---
title: "Final Project"
author: "Bryan Winmill - winmill2 AND Justin Cherry - wcherry2"
date: "7/17/2019"
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---


# Project Proposal

### Team Members - Net Id
- Bryan Winmill - winmill2 - This is a test
- Justin Cherry - wcherry2

### Tentative Project Title:
- College Basketball School Performance

### Description of Dataset:
This dataset includes data from 479 College basketball teams that were classified as a Division I school (or equivalent) for at least one season from 1893 to 2019. The following categories (predictors) are from their website: 

- Rank: Rank of School
- School: Name of School
- City: School location **(categorical predictor)**
- State: School location **(categorical predictor)**
- From: First year classified as Division I **(categorical predictor)**
- To: Last year classified as Division I **(categorical predictor)**
- Yrs: Number of years played
- G: Total Games played
- W: Total Games won
- L: Total Games lost
- W-L% Win-Loss percentage **(This will be our response variable)**
- SRS: (Simple Rating System): A rating that takes into account average point differential and strength of schedule. The rating is denominated in points above/below average, where zero is average. Non-Division I games are excluded from the ratings
- SOS: (Strength of Schedule): A rating of strength of schedule. The rating is denominated in points above/below average, where zero is average. Non-Division I games are excluded from the ratings.
- AP:  Number of schools ranked in final AP poll (poll taken before the conclusion of the NCAA tournament)
- CREG: Number of regular season conference championships won
- CTRN: Number of conference tournament championships won
- NCAA: Number of NCAA Tournament appearances
- FF: Number of NCAA Final Four appearances
- NC: Number of NCAA Tournament championships won

### Background information on dataset:
- https://www.sports-reference.com/cbb/schools/
- The data shows schools and the number of seasons they have played, their wins, losses, win-loss ratio, SRS, SOS, and then how many times they appeared in certain championships, and whether they made it to the final 4, or if they won the championship.

### Statement of personal interest:
- We thought it would be interesting to see if you can make a semi-accurate prediction of a college schools performance, specifically Win-Loss percentage and Championship wins, based on some of the other given data in the dataset (Such as years/seasons played, total games played, Simple Rating System, and Strength Of Schedule).


### Evidence the data can be loaded into R.


```{r}
library(readr)
school_data = read_csv("SchoolData.csv")
```

```{r}
nrow(school_data)
```

```{r}
head(school_data)
```

